Lab 2 Summary Statistics for Multivariate Data
========================================================

SOLUTION
===

```{r readdata, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
pisa <- read.csv("PISA-USA.csv", stringsAsFactors=FALSE)
dim(pisa)
colnames(pisa)
pisamath <- pisa[, 1:6]
```

```{r summaries, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
summary(pisamath)
library(matrixStats)
colMeans(pisamath[,-1])
colSds(as.matrix(pisamath[,-1]))
var(pisamath[,-1])
cor(pisamath[,-1])
```

### Exercise 1 - Focusing on the 5 overall math scores only.

1.What is the sample size, *n*? 10294

2.What is the dimension of the data, *p*? 5

3.Tabulate the means and standard deviations.

 -- | PV1MATH | PV2MATH | PV3MATH | PV4MATH | PV5MATH |
---|---:|---:|---:|---:|---:|
Mean | 485.6  | 485.8  | 485.9 |  486.5 |  485.7 |
SD   | 92.4| 93.1  | 92.7 |  92.6 |  92.9 |

4.Compute and report the correlation matrix.

Two digits is enough to report correlation!

 --|        PV1MATH | PV2MATH|  PV3MATH | PV4MATH | PV5MATH |
 ---|---:|---:|---:|---:|---:|
 PV1MATH | 1.00 | 0.93 | 0.93 | 0.93 | 0.93 |
 PV2MATH | 0.93 | 1.00 | 0.93 | 0.93 | 0.93 |
 PV3MATH | 0.93 | 0.93 | 1.00 | 0.93 | 0.93|
 PV4MATH | 0.93 | 0.93 | 0.93 | 1.00 | 0.93 |
 PV5MATH | 0.93 | 0.93 | 0.93 | 0.93 | 1.00 |
 
5.Summarize in a paragraph what you learn about the US math scores based on the sumamry statistics.

The averages and standard deviations for the five different math scores are very similar, and the correlation between the scores are very strong. This would indicate that if we know one, we know all five. It would be appropriate to work with the average of the five scores.

```{r gendersummaries, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
library(dplyr)
summarize(group_by(pisamath, Gender), 
          m1=mean(PV1MATH), m2=mean(PV2MATH), m3=mean(PV3MATH), m4=mean(PV4MATH), m5=mean(PV5MATH),
          s1=sd(PV1MATH), s2=sd(PV2MATH), s3=sd(PV3MATH), s4=sd(PV4MATH), s5=sd(PV5MATH))
head(pisamath)
cor(as.matrix(filter(pisamath, Gender == "Female")[,-1]))
cor(as.matrix(filter(pisamath, Gender == "Male")[,-1]))
```

### Exercise 2 - Gender in relation to math scores.

1.Compute summary statistics by gender (mean, sd, correlation).

 -- | Gender | PV1MATH | PV2MATH | PV3MATH | PV4MATH | PV5MATH |
---|---|---:|---:|---:|---:|---:|
Mean | Female | 480.4 | 481.0 | 480.9 | 481.6 | 480.6|
Mean | Male | 490.8 | 490.7 | 490.8 |491.4 | 490.8|
SD  |Female | 90.32 | 91.22 | 90.44 | 90.57 | 90.42 |
SD  | Male |94.25 |94.47 | 94.71 | 94.33 | 95.08 |

Females:

 --|        PV1MATH | PV2MATH|  PV3MATH | PV4MATH | PV5MATH |
 ---|---:|---:|---:|---:|---:|
 PV1MATH | 1.00 | 0.92 | 0.92 | 0.92 | 0.92 |
 PV2MATH | 0.92 | 1.00 | 0.92 | 0.93 | 0.92 |
 PV3MATH | 0.92 | 0.92 | 1.00 | 0.92 | 0.92|
 PV4MATH | 0.92 | 0.93 | 0.93 | 1.00 | 0.93 |
 PV5MATH | 0.92 | 0.92 | 0.92 | 0.93 | 1.00 |

Males:

 --|        PV1MATH | PV2MATH|  PV3MATH | PV4MATH | PV5MATH |
 ---|---:|---:|---:|---:|---:|
 PV1MATH | 1.00 | 0.93 | 0.93 | 0.93 | 0.93 |
 PV2MATH | 0.93 | 1.00 | 0.93 | 0.93 | 0.93 |
 PV3MATH | 0.93 | 0.93 | 1.00 | 0.93 | 0.93|
 PV4MATH | 0.93 | 0.93 | 0.93 | 1.00 | 0.93 |
 PV5MATH | 0.93 | 0.93 | 0.93 | 0.93 | 1.00 |
 
2.Write a summary comparing and contrasting male and female math scores.

The only differnce between males and females is that the average females scores are approximately 10 points lower than the average male scores. Perhaps variation in female scores is slightly less. Correlation between scores is the same. 

```{r mathreadsci, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
pisa.all <- mutate(pisa, math = apply(cbind(PV1MATH, PV2MATH, PV3MATH, PV4MATH, PV5MATH), 1, mean),
          read = apply(cbind(PV1READ, PV2READ, PV3READ, PV4READ, PV5READ), 1, mean), 
          science = apply(cbind(PV1SCIE, PV2SCIE, PV3SCIE, PV4SCIE, PV5SCIE), 1, mean))
dim(pisa.all)
pisa.mrs <- select(pisa.all, c(Gender, math:science))
dim(pisa.mrs)
summarize(group_by(pisa.mrs, Gender), 
          mmath=mean(math), mread=mean(read), mscience=mean(science),
          smath=sd(math), sread=sd(read), sscience=sd(science))
cor(as.matrix(filter(pisa.mrs, Gender == "Female")[,-1]))
cor(as.matrix(filter(pisa.mrs, Gender == "Male")[,-1]))
```

### Exercise 3 - Combining values, and exploring reading and science scores

1.Average the 5 scores for each type of test, to get a single mean for each student (Nothing to report here)

2.Compute summary statistics by gender

 -- | Gender | Math | Read | Science 
---|---|---:|---:|---:|
Mean | Female | 480.9 | 516.6 |   499.4 |
Mean | Male | 490.9 | 489.8  |  504.4 |
SD  |Female | 87.79 | 86.24 |   88.71 | 
SD  | Male |91.97 | 92.12  |  95.16 |

Females:

 --|        Math | Read | Science |
 ---|---:|---:|---:|---:|---:|
 Math | 1.00 | 0.91 | 0.93 | 
 Read | 0.91 | 1.00 | 0.93 | 
 Science | 0.93 | 0.93 | 1.00 | 
 
Males:

  --|        Math | Read | Science |
 ---|---:|---:|---:|---:|---:|
 Math | 1.00 | 0.92 | 0.94 | 
 Read | 0.92 | 1.00 | 0.93 | 
 Science | 0.94 | 0.93 | 1.00 | 
 
3.Write a summary comparing and contrasting male and female scores.

Although on average boys do better than girls on math by 10 points, girls do better than boys on reading by 25 points. Scores are closer on science. All three sets of scores are strongly correlated for both boys and girls. 

```{r math, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
pisa.all <- mutate(pisa, acc = apply(cbind(PV1MACC, PV2MACC, PV3MACC, PV4MACC, PV5MACC), 1, mean),
          acq = apply(cbind(PV1MACQ, PV2MACQ, PV3MACQ, PV4MACQ, PV5MACQ), 1, mean), 
          acs = apply(cbind(PV1MACS, PV2MACS, PV3MACS, PV4MACS, PV5MACS), 1, mean),
          acu = apply(cbind(PV1MACU, PV2MACU, PV3MACU, PV4MACU, PV5MACU), 1, mean),
          ape = apply(cbind(PV1MAPE, PV2MAPE, PV3MAPE, PV4MAPE, PV5MAPE), 1, mean),
          apf = apply(cbind(PV1MAPF, PV2MAPF, PV3MAPF, PV4MAPF, PV5MAPF), 1, mean),
          api = apply(cbind(PV1MAPI, PV2MAPI, PV3MAPI, PV4MAPI, PV5MAPI), 1, mean))
dim(pisa.all)
pisa.allmath <- select(pisa.all, c(Gender, acc:api))
dim(pisa.allmath)
summary(pisa.allmath)
pisa.allmath.means <- summarize(group_by(pisa.allmath, Gender), 
          macc=mean(acc, na.rm=T), macq=mean(acq, na.rm=T), macs=mean(acs, na.rm=T),
          macu=mean(acu, na.rm=T), mape=mean(ape, na.rm=T), mapf=mean(apf, na.rm=T), mapi=mean(api, na.rm=T),
          sacc=sd(acc, na.rm=T), sacq=sd(acq, na.rm=T), sacs=sd(acs, na.rm=T), sacu=sd(acu, na.rm=T), 
          sape=sd(ape, na.rm=T), sapf=sd(apf, na.rm=T), sapi=sd(api, na.rm=T))
pisa.allmath.means[,1:7]
pisa.allmath.means[,8:14]
cor(as.matrix(filter(pisa.allmath, Gender == "Female")[,-1]), use="complete.obs")
cor(as.matrix(filter(pisa.allmath, Gender == "Male")[,-1]), use="complete.obs")
```

### Exercise 4 - Math scores

1.How many missing values are there? Ho wmany variables have missing values? What proportion of cases are missing on each variable?

There are 4978 missing values, the same number on each of the seven variables. This corresponds to 48% missing in the data.

2.Compute summary statistics by gender

 -- | Gender | ACC | ACQ | ACS | ACU | APE | APF | API |
---|---|---:|---:|---:|---:|---:|---:|---:|
Mean | Female | 488.2 | 476.7| 464.6| 491.8 |482.8 |475.5|491.3|
Mean | Male | 506.6 | 489.4 | 477.8 | 503.5 | 492.4 | 494.2| 506.3|
SD  |Female | 96.16 | 98.54 | 97.7 |89.22| 89.29 | 99.98 | 95.41|
SD  | Male |101.49 |101.40 |101.9| 93.53| 93.03| 104.98 |101.35|

Females:

 --|        ACC | ACQ | ACS | ACU | APE | APF | API |
 ---|---:|---:|---:|---:|---:|---:|---:|
ACC | 1.00 | 0.90 | 0.90 | 0.90 | 0.93 | 0.92 | 0.92 |
ACQ | 0.90 | 1.00 | 0.90 | 0.91 | 0.95 | 0.92 | 0.92 |
ACS | 0.90 | 0.90 | 1.00 | 0.90 | 0.93| 0.93 | 0.89 |
ACU | 0.90 | 0.91 | 0.90 | 1.00 | 0.93 |0.92| 0.93|
APE | 0.92 | 0.95 | 0.93 | 0.93 | 1.00 |0.93|0.93|
APF | 0.92 | 0.92 | 0.93 | 0.92 | 0.93 |1.00|0.92|
API | 0.92 | 0.92 | 0.89 | 0.93 | 0.93 |0.92|1.00|
 
Males:

 --|        ACC | ACQ | ACS | ACU | APE | APF | API |
 ---|---:|---:|---:|---:|---:|---:|---:|
ACC | 1.00 | 0.92 | 0.91 | 0.91 | 0.94 | 0.93 | 0.93 |
ACQ | 0.92 | 1.00 | 0.91 | 0.92 | 0.95 | 0.92 | 0.92 |
ACS | 0.91 | 0.91 | 1.00 | 0.90 | 0.93| 0.93 | 0.90 |
ACU | 0.91 | 0.92 | 0.90 | 1.00 | 0.93 |0.92| 0.93|
APE | 0.94 | 0.95 | 0.93 | 0.93 | 1.00 |0.93|0.93|
APF | 0.93 | 0.92 | 0.93 | 0.92 | 0.93 |1.00|0.92|
API | 0.93 | 0.92 | 0.90 | 0.93 | 0.93 |0.92|1.00|

3.Write a summary comparing and contrasting male and female scores.

Across the board of all of the different types of math skills boys score better than girls on average, in the USA. The difference varies a little from 10-20 points. The variation in girls scores is slightly smaller across the board. The correlation is uniformly strong both both genders.
