Lab 3 Plotting Multivariate Data
========================================================

# SOLUTION

### Music

The music data, music-plusnew-sub.csv. The data has 5 quantitative variables, where first two columns (Artist, Type)
describe the music piece. The raw data are time series for each music clip. On each time series the average amplitude, variance of amplitude, maximum amplitude, and two variables calculated on the spectral decomposition are calculated. The tracks are
typed as either Rock, Classical or New Wave. And there are 5 tracks that are not identified.

```{r readmusic}
music <- read.csv("music-plusnew-sub.csv", row.names=1, stringsAsFactors=FALSE)
dim(music)
colnames(music)
head(music)
summary(music)
table(music$Type)
table(music$Artist)
```

### Bodyfat

The body fat data bodyfat.csv contains various measurements on 252 (normal) males.

```{r readbodyfat}
bodyfat <- read.csv("bodyfat.csv")
dim(bodyfat)
colnames(bodyfat)
head(bodyfat)
```

### PISA

Data set studies in the first lab on test scores for 15 year olds. Only the USA measurements are examined. These are the scores for the different types of math skills. 

```{r readpisa}
pisamath <- read.csv("pisamathmeans.csv")
dim(pisamath)
summary(pisamath)
```

### Challenge data set

The clusters-unknown.csv data has simulated data with number of known, but unrevealed, clusters in it. It is available from  http://ggobi.org/book/data/clusters-unknown.csv.

```{r readclusters}
cl <- read.csv("http://ggobi.org/book/data/clusters-unknown.csv")
dim(cl)
head(cl)
```

# Software

The ggplot2 package in R, the GGally package and ggobi. The main web page for ggplot2 is http://had.co.nz/ggplot2, but this page http://www.cookbook-r.com/Graphs/ has really good practical examples. Only the man pages exist for help for the GGally package. For ggobi look at all the documentation at http://www.ggobi.org.

```{r loadlibs}
library(ggplot2)
library(GGally)
```

# Exercise 1 Music clips data: Using ggplot2.

```{r musicplots, fig.width=7, fig.height=7}
table(music$Type)
music.sub <- subset(music, Type != "New wave")
qplot(LAve, data=music.sub, geom="histogram") + facet_wrap(~Type, ncol=1)
qplot(LVar, LAve, data=music.sub, color=Type, size=I(5), alpha=I(0.5)) + theme(aspect.ratio=1)
```

1.Subset the data so that only the major music types (Rock and Classical) are present. (Nothing to hand in.)
2.Make histograms of LAve facetted by Type. Set the binwidth to units of 10. How does the average value vary by type of music? 

*Classical music has a right skewed distribution. Rock music has a bimodal distribution with two clusters, and perhaps one outlier. Classical music has generally higher averages than Rock music.*

3. Make a scatterplot of LVar vs LAve, with points colored by the
  type of music. Explain how the two major groups (Rock and Classical) differ.

*Rock generally has low averages, but high variances. It has a subgroup with really low averages, and one outlier with a very high variance. On the other hand Classical music has higher averages but lower variances, and there is one outlier with extremely high average.*

# Exercise 2 Body fat data: Using GGally.

```{r plotbf, fig.width=8, fig.height=8}
ggpairs(bodyfat[,1:4])
bf.sub <- subset(bodyfat, Heightinches > 60 & Weightlbs < 300)
ggpairs(bf.sub[,1:4])
```

```{r plotbf2, fig.width=12, fig.height=8}
ggparcoord(bf.sub, 5:14)
summary(bf.sub$AnkleCircm)
bf.sub[bf.sub$AnkleCircm > 30,]
bf.sub2 <- subset(bf.sub, AnkleCircm < 30)
ggparcoord(bf.sub2, 5:14)
```

1.Make a scatterplot matrix of the first four variables, \% Body Fat, Age, Weight and Height. There's a problem with the data. What is it?

*There is one male who is unbelievably short! There is also one male who is quite heavy but it is possible.*

2.Fix the problem with the data and remake the scatterplot matrix. Explain the association between the variables.

*Now the data looks very regular.  \% body fat is strongly positively associated with Weight. Weight and height are positively associated. The other variables have weak to no association.*

3.Make a parallel coordinate plot of the last 10 variables, the circumference variables. There are some outliers in these measurements. Explain the ones that you see.

*There is one case which has an extremely high value for the first 8 variables. There are two cases that have unusually high values for AnkleCircm.*

4. Remove the outliers and remake the parallel coordinate plot. Describe the structure of the data.

*There's no particularly strong structure left. The measurements bounce around the median value on each variable, some men have larger neck circumference but relatively lower chest circumference. One man has relatively large values on several of the variables. Forearm circumference is a little odd: there are two cases with really low values, but these men have relatively large bicep values. And there is one case with the reverse pattern.  This data is a reasonably good candidate for modeling using a multivariate normal distribution.*

# Exercise 3 PISA Maths scores

```{r plotpisa, fig.width=12, fig.height=12}
pisamath <- subset(pisamath, !is.na(acc))
summary(pisamath)
ggpairs(pisamath, colour="Gender", alpha=I(0.5))
```

1. How many missing values for variable acc? Remove the missing values in the data. *4978*
2. Make a scatterplot matrix of the variables, with points coloured by Gender. Describe the association between the variables.

*There is very strong association between the different tpes of math scores. The slight shift in average scores is possible to see, ever so slightly in the histograms. Males and females overlap more than they are different. Even though there is a 5-15 point difference on average, performances are very similar, with many girls doing better than many boys. The average differences are very small relative to the overall variation in scores.*

# Exercise 4 Using tours in ggobi

This R code will not work well with knitr, because it opens software to do interactive graphics. You should run the lines of code in the command line, and then work with the graphical user interface of ggobi to answer the questions. (WARNING: rggobi will crash your R session from time to time, so be sure to save your .Rmd file frequently. )

```
library(rggobi)
rggobi(music)
```

1.Open the music data in ggobi. Color the types of music differently. Is it possible to completely distinguish between the types of music based on these 5 variables? 

*Not quite but close. Rock and classical have mostly different values, forming two separate clusters in the 5D space, but the clusters are not separated. There is a little bit of overlap between them, some Rock tracks are like some Classical tracks and vice versa. The new wave tracks are similar to both other music types, falling in the middle roughly.*

2.Find and name 2-3 unusual, potentially outlying, music tracks.

*V6, V2, V7, Saturday Morning all are outliers in some projections.*

3.What's strange about the Abba tracks?

*The Abba tracks form their own cluster, and differ from the other Rock tracks. The difference is primarily due to having very low values on LAve.*

# Exercise 5 (EXTRA CREDIT) Cluster challenge data

```
rggobi(cl)
```

Use the brush-and-spin approach to decide how many clusters in the data. *4*


