---
title: "Introduction"
author: "Di Cook"
date: "March 16, 2015"
output: 
  ioslides_presentation:
    css: styles.css
    mathjax: local
    self_contained: false
---

## What is business analytics?

Business analytics involves using exploring data to obtain new insights and understanding of business performance. It involves exploratory data analysis, visualization and statistical methods.

It may be interpreted more broadly than business intelligence (data mining) which focuses more on metrics for describing and predicting performance.

## Components of business analytics

- Pulling together and cleaning data
- Exploring and visualizing data
- Fitting, comparing and assessing models
- General tools for understanding randomness: resampling, permutation, cross-validation
- General tools for fitting models: optimization, training and testing

## Pulling data together 1a

Historical exchange rates extracted from [http://www.oanda.com/currency/historical-rates/](http://www.oanda.com/currency/historical-rates/).

```{r xc1}
xc <- read.csv("../data/exchangerate.csv", stringsAsFactors=FALSE)
dim(xc)
head(xc)
```

## Pulling data together 1b

```{r xc2, message=FALSE}
xc$Date <- as.Date(paste("1",xc$Date,sep="-"), "%d-%b-%y")
summary(xc)
```

## Pulling data together 2a

Historical GDP extracted from [http://www.abs.gov.au/AUSSTATS](http://www.abs.gov.au/AUSSTATS/abs@.nsf/DetailsPage/5206.0Dec%202014?OpenDocument).

```{r gdp1, message=FALSE}
gdp <- read.csv("../data/gdp.csv", stringsAsFactors=FALSE)
dim(gdp)
head(gdp)
```

## Pulling data together 2b

```{r gdp2, message=FALSE}
gdp$Date <- as.Date(paste("1", gdp$Date,sep="-"), "%d-%b-%Y")
summary(gdp)
```

## Merge data 1

```{r xc-gdp1, message=FALSE}
library(dplyr)
gdp <- filter(gdp, Date >= min(xc$Date))
summary(gdp)
```

## Merge data 2

```{r xc-gdp2, message=FALSE}
xc.gdp <- left_join(xc, gdp)
dim(xc.gdp)
head(xc.gdp)
```

## Plotting data 1

```{r plotAUD, message=FALSE}
library(ggplot2)
qplot(Date, AUD, data=xc.gdp, geom="line")
```

## Plotting data 1

```{r plotGDP, message=FALSE, warning=FALSE}
qplot(AUD, GDP.per.capita, data=xc.gdp) + theme(aspect.ratio=1)
```

## Equations

Equations can be written using latex notation, e.g. 

Variables in data are usually denoted with $X$, and response variables as $Y$. Predicted response variables are $\hat{Y}$. 